Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-14 paper/paper.tex
Updated section on liveness analysis for lazy languages
Updated section on liveness analysis

\subsection{Liveness Analysis}

@@

The concept of demand and liveness for lazy languages are almost similar to their 
counterparts in the eager language. The important distinction comes from the fact 
that in eager languages, variables can point only to data values (either atomic values or cons cells). 
Hence any demand on the variable can easily be transferred onto the data being pointed. 
In lazy languages, a variable can either point to a fully evaluated data value or it 
could point to a closure which on evaluation generates the required data value. For example, 
consider the statement {\tt let \py~ $\leftarrow $ (\CDR~\px)}. In case of an eager language 
the variable \py~ points to  the \CDR~ part of \px~ which is guaranteed to be a cons cell. 
Hence any demand on \py~ can be immediately transferred to the \CDR~ part of \px. In case of 
a lazy language, \px~ might not be evaluated and  hence pointing to a closure. In our 
example, it points to the closure corresponding to (\CONS~1~\px). It does not make any sense 
to transfer the demand onto the \CDR~ part of \px~ as it still does not exist. In such a case 
we treat the closure as a demand transformer and tranform the demand on \py~ to the arguments 
of the closure pointed by \px. This procedure is similar to the process of transferring the demand 
from the return value of a function to the arguments of the function call in an eager language. In case 
of lazy languages this process cannot be done statically as we cannot statically determine whether a variable points to 
a data value or a closure. Our method approximates the liveness of closure arguments at runtime 
by tagging extra information with each closure. Since the arguments of the closures could themselves be closures in turn, 
we keep recursively applying the same procedure until we hit a data value. The only case where we do not 
follow this procedure is if the demand on a particular closure is null. In such a case we safely collect 
the entire closure as garbage.

@@

  In  the  sequel we  shall  describe  how  to obtain  $\cal{M}$.  The
  formulation  of $\cal{P}$  is easy  and we  do not  describe  it any
  further.

>>>>

\subsection{Liveness Analysis for lazy languages}

@@

The main challenges in doing liveness analysis for lazy languages are, 
1) a variable can either point to a fully evaluated data value or it 
could point to a closure which on evaluation generates the required data value. For example, 
consider the statement {\tt let \py~ $\leftarrow $ (\CDR~\px)}. In case of an eager language 
the variable \py~ points to  the \CDR~ part of \px~ which is guaranteed to be a cons cell. 
Hence any demand on \py~ can be immediately transferred to the \CDR~ part of \px. In case of 
a lazy language, \px~ might not be evaluated and  hence pointing to a closure. In our 
example, it points to the closure corresponding to (\CONS~1~\px). It does not make any sense 
to transfer the demand onto the \CDR~ part of \px~ as it still does not exist.
, 2) it is not possible to statically determine when a closure gets evaluated.  Therefore 
during GC we treat a closure as a demand transformer and tranform the demand on to the arguments 
of the closure pointed. This is similar to the process of transferring the demand 
from the return value of a function to the arguments of the function call in an eager language.

In our example, during GC if we find that \px~ points to a closure instead of a cons cell we 
find out the demand/liveness on the arguments of the closure, namely \py.
Statically it cannot be determined what values a closure points to or when it gets evaluated,
hence our method approximates the liveness of closure arguments at runtime 
by tagging extra information with each closure. Since the arguments of the closures could themselves be closures in turn, 
we recursively apply the same procedure until we hit a data value. The only case where we do not 
follow this procedure is if the demand on a particular closure is null. In such a case we safely collect 
the entire closure as garbage.

@@

  In  the  sequel we  shall  describe  how  to obtain  $\cal{M}$.  The
  formulation  of $\cal{P}$  is easy  and we  do not  describe  it any
  further.
  We observe that the ${ref}$ rules neither use ${\cal M}$ directly nor 
  create new labels. We only need to consider the rules for {\LET}, {\SIF}, and
  {\RETURN}.

  A {\RETURN~\px}\  statement always forces an evaluation and hence a new 
  label {\em l} needs to be created and added to $\cal{M}$. A new liveness
  environment is associated with {\em l}. The demand $\sigma$ that is placed 
  on this expression is now transferred to the variable \px. {\em l} will now
  contain an entry for \px~ with liveness $\sigma$.

  An \SIF~ statement will modify  $\cal{M}$ by doing a union of 
  the maps from the {\em then} and the {\em else} part. Note that each part can contain 
  more than one label. Since the variable corresponding to the condition may not
  be part of either the {\em then} or the {\em else} part, a new label {\em l} is created 
  and added to $\cal{M}$. The only liveness associated with {\em l} is the liveness 
  of the condition variable.

  The processing rule for {\LET} is the most complex among all the rules. A {\LET} does
  not add any new label but might modify the liveness values of some variables or might add 
  liveness values for the let bound variable. Since there might be more than one {\RETURN} 
  statement following a {\LET}, there might be multiple {\em E-paths} reaching the {\LET}.
  The demand on the let variable might be different along different {\em E-paths}. Therefore,
  for each label {\em l} in $\cal{M}$, the liveness of the let-expression is calculated using 
  the liveness environment $\Lv$ corresponding to {\em l}. Since the {\LET} is lazy, 
  calculating its livness might require a fixed-point processing.
